---
title: 'Revisiting the concept of the "Neolithic Founder Crops" in southwest Asia'
subtitle: "Data analysis"
author:
  - first_name: "Joe"
    last_name: "Roe"
    url: https://joeroe.io
    orcid_id: 0000-0002-1011-1244
  - first_name: "Amaia"
    last_name: "Arranz-Otaegui"
    orcid_id: 0000-0002-5091-6426
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 4
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE
)
```

```{r deps}
library("tidyverse")
library("SWAsiaNeolithicFounderCrops")
library("swapdata")
library("sf")
library("ggspatial")
library("ggstream")
library("here")
library("gt")
library("glue")
library("patchwork")
library("khroma")
```

This document describes the quantitative analysis our paper *Revisiting the concept of the “Neolithic Founder Crops” in southwest Asia* (Arranz-Otaegui & Roe 2022).
A research compendium containing the data and R code needed to reproduce the analysis can be accessed at <https://github.com/joeroe/SWAsiaNeolithicFounderCrops>).

TODO: instructions for interactive use? (Binder?)

## Methods & Materials

### Study region and period

```{r data-spatial, fig.cap='The study region', results='hide', layout='l-body-outset'}
proj <- 22770
unproj <- 4326

sw_asia <- st_read(here("analysis/data/raw_data/regions/swasia.geojson"))

land <- st_read(here("analysis/data/raw_data/ne/ne_50m_land.shp"))
rivers <- st_read(here("analysis/data/raw_data/ne/ne_50m_rivers_lake_centerlines_scale_rank.shp"))
lakes <- st_read(here("analysis/data/raw_data/ne/ne_50m_lakes.shp"))
highland <- st_read(here("analysis/data/raw_data/srtm/srtm30_sw_asia_1000m.geojson"))
highland <- filter(highland, above_1000m == 1)

base_map <- ggplot() +
  shadow_spatial(st_bbox(sw_asia), fill = NA) +
  annotation_spatial(land, fill = "white") +
  annotation_spatial(highland, fill = "#f0f0f0", colour = NA) +
  annotation_spatial(rivers, colour = "#555555", size = 0.5) +
  annotation_spatial(lakes, fill = "lightblue") +
  coord_sf(expand = FALSE, crs = proj)

base_map
```

We aimed to collate archaeobotanical data from Southwest Asia (fig. \@ref(fig:data-spatial)) from the Neolithic period, c. 11,700 – 6500 BP (table \@ref(tab:periods)).
To mitigate against an "edge effect", we also included samples in the source databases dated to the preceding Late Epipalaeolithic (c. 15,000–11,700 BP) and succeeding Chalcolithic (c. 6500–5000 BP).

```{r periods}
periods <- tribble(
  ~period,             ~agriculture,                    ~start_bp, ~end_bp,
  "Late Epipal.",      "Foraging",                      15000,     11700,
  "PPNA",              "Pre-domestication cultivation", 11700,     10700,
  "EPPNB",             "Domestication",                 10700,     10200,
  "MPPNB",             "Domestication",                 10200,     9500,
  "LPPNB/C",           "Agriculture",                   9500,      8500,
  "Pottery Neolithic", "Agriculture",                   8500,      6500,
  "Chalcolithic",      "Agriculture",                   6500,      5000,
)

neolithic <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C", "Pottery Neolithic")

periods %>% 
  gt(caption = "Cultural periods used in the analysis") %>% 
  cols_label(period = "Period", agriculture = "Agricultural stage",
             start_bp = "Start (cal BP)", end_bp = "End (cal BP")
```

### Archaeobotanical data

We collated data from three large-scale archaeobotanical databases: ORIGINS [@Wallace2018], and COMPAG [@Lucas2018; @Fuller2018], and ADEMNES [@Riehl2005].

The process of importing, cleaning, and making these databases comparable (see R functions in `R/data.R` for full details) was fairly involved.
In short, we:

* Standardised site names across databases using the thesaurus compiled at <https://github.com/joeroe/swapdata>
* Standardised taxonomic names across databases and applied our additional classifications (described in `analysis/data/raw_data/founder_crops.tsv`)
* When sites were in multiple databases, preferred records from the most detailed/recent one (i.e. ORIGINS > COMPAG > ADEMNES)
* ORIGINS is recorded by sample, but COMPAG and ADEMNES by phase, so we aggregated ORIGINS by phase for comparability (and because individual samples are more likely to be 'noisy')
* Calculated phase-level proportions for COMPAG

We filtered this data to include only samples from our region (Southwest Asia) and period (11,700–6500 BP) of interest,
with a buffer of ±2,000 years to reduce "edge effects" in our time series analyses.

```{r data-flora}
start_bp <- first(periods$start_bp)
end_bp <- last(periods$end_bp)
flora <- collate_flora(
  origins_path = here("analysis/data/raw_data/origins/"),
  ademnes_path = here("analysis/data/raw_data/ademnes"),
  compag_path = here("analysis/data/raw_data/compag2018"),
  taxa_path = here("analysis/data/raw_data/founder_crops.tsv"),
  region = sw_asia,
  start_bp = start_bp,
  end_bp = end_bp
)

# Summary totals
flora %>% 
  distinct(site_name) %>% 
  nrow() ->
  n_site

flora %>% 
  distinct(site_name, phase_code) %>% 
  nrow() ->
  n_assemb
```

The resulting collated dataset (`analysis/data/derived_data/swasia_neolithic_flora.tsv`) includes archaeobotanical assemblages representing `r n_assemb` distinct phases from `r n_site` sites.

### Chronological data

All three databases include rough absolute date ranges (in the vast majority of cases summarised the available radiocarbon dates) for each sample.
Using this data, each phase was classified into one of the cultural periods defined above using its mid-point date (table \@ref(tab:sample-periods)).

```{r sample-periods}
flora <- mutate(
  flora,
  age_mid = age_end + ((age_start - age_end) / 2),
  period = cut(-age_mid, 
               breaks = -c(periods$start_bp, 0), 
               labels = periods$period,
               ordered_result = TRUE)
)

flora %>% 
  distinct(period, site_name, phase) %>% 
  add_count(period, name ="n_phases") %>% 
  distinct(period, site_name, n_phases) %>% 
  add_count(period, name = "n_sites") %>% 
  distinct(period, n_phases, n_sites) %>% 
  select(period, n_sites, n_phases) %>% 
  arrange(period) %>%
  gt(caption = "Sites and assemblages per period") %>% 
  cols_label(period = "Period", n_sites = "Sites", n_phases = "Assemblages") 

flora %>% 
  distinct(site_name, phase_code, period) %>% 
  filter(period %in% neolithic) %>% 
  nrow() ->
  n_assemb_neolithic

flora %>% 
  filter(period %in% neolithic) %>% 
  distinct(site_name) %>% 
  nrow() ->
  n_site_neolithic
```

For a finer-grained chronology, we also sliced the assemblages into century bins.
An assemblage was considered to belong to a bin (e.g. 5000–5099 BP) if any part of its absolute date range falls within that bin – this results in the duplication of data across bins, but better reflects the inherent imprecision of radiocarbon dating.

```{r sample-centuries}
centuries <- seq(from = first(periods$start_bp),
                 to = last(periods$end_bp),
                 by = -100)
names(centuries) <- centuries

flora %>% 
  mutate(century = map_dfc(centuries, ~. < age_start & . > age_end)) %>% 
  unpack(century) %>% 
  pivot_longer((ncol(flora)+1):ncol(.), names_to = "century",
               names_transform = list(century = as.integer)) %>% 
  filter(value) %>% 
  select(-value) %>% 
  mutate(period = cut(-century, 
               breaks = -c(periods$start_bp, last(periods$end_bp)), 
               labels = periods$period,
               ordered_result = TRUE)) ->
  flora_ts
```

### Sample coverage

We are now in a position to inspect the geographical (fig. \@ref(fig:sample-map)) and temporal \@ref(fig:sample-chronology) coverage of the data.
Unsurprisingly, the distribution is uneven. 
Notably, we have few Epipalaeolithic sites, but a very large number of Chalcolithic sites;
the regional coverage reflects broader trends in the research history of the region, with a large body of evidence from the Southern Levant, especially for earlier periods, and the rest of the region more patchily covered.
Nevertheless, we have a decent baseline number of samples for each time slice in our analysis—the lowest is 11 assemblages, for the EPPNB—especially for the key Neolithic periods.

```{r sample-summary}
flora %>% 
  group_by(site_name, phase_code) %>% 
  summarise(across(c(source, latitude, longitude, phase, age_start, age_end,
                     period), first),
            .groups = "drop") ->
  flora_phases
```


```{r sample-map, fig.cap='Geographical distribution of sampled archaeobotanical assemblages', layout='l-body-outset'}
p1 <- base_map + 
  geom_spatial_point(data = flora_phases, 
                     mapping = aes(longitude, latitude),
                     size = 0.5) +
  labs(x = NULL, y = NULL)
  
p1_left <- p1 + facet_wrap(vars("All sites"))
p1_right <- p1 + 
  facet_wrap(vars(period)) + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank())

p1_left + p1_right

# ggsave(here("analysis/figures/sample-map.pdf"), width = 235, height = 160, 
#        units = "mm")
# ggsave(here("analysis/figures/sample-map.png"), width = 160, height = 80,
#        units = "mm", dpi = 300)
```


```{r sample-chronology, fig.cap='Chronological distribution of sampled archaeobotanical assemblages', layout='l-body-outset'}
p2_top <- flora_phases %>% 
  count(period) %>% 
  left_join(periods) %>% 
  mutate(period = ordered(period, levels = periods$period)) %>% 
  ggplot(aes(xmin = start_bp, xmax = end_bp, ymin = 0, ymax = n,
             fill = period)) +
  geom_rect() +
  scale_x_reverse(limits = c(15000, 5000),
                  breaks = seq(15000, 5000, -1000)) +
  scale_y_continuous(limits = c(0, 150)) +
  khroma::scale_fill_muted() +
  labs(x = NULL, y = "Samples by period", fill = "Period") +
  theme_minimal() +
  theme(legend.position = "top", 
        axis.text.x = element_blank(), axis.ticks.x = element_blank())

p2_bottom <- flora_ts %>% 
  group_by(site_name, phase_code, century) %>% 
  summarise(period = first(period), .groups = "drop") %>% 
  ggplot(aes(century, fill = period)) +
  geom_bar(width = 100) +
  scale_x_reverse(limits = c(15000, 5000), 
                  breaks = seq(15000, 5000, -1000)) +
  scale_y_reverse(limits = c(150, 0)) +
  khroma::scale_fill_muted(guide = guide_none()) +
  labs(x = "cal BP", y = "Samples by century") +
  theme_minimal() +
  theme(legend.position = "bottom")

p2_top / p2_bottom

# ggsave(here("analysis/figures/sample-chronology.pdf"), width = 160, height = 100, 
#        units = "mm")
```

## Results

### Were the founder crops the *most important* plants?

#### Table of occurrences

The following table summarises the occurrence of each standardised taxon in Neolithic assemblages, broken down by plant-category.

```{r occ-all, layout='l-body-outset'}
flora %>% 
  drop_na(category, taxon) %>% 
  filter(period %in% neolithic) %>% 
  group_by(category, taxon, phase_code) %>% 
  summarise(prop = sum(prop, na.rm = TRUE), .groups = "drop_last") %>% 
  summarise(
    n_present = sum(prop > 0),
    n_gthalf = sum(prop > 0.5),
    n_gtqrtr = sum(prop > 0.25),
    .groups = "drop_last"
  ) %>% 
  mutate(
    pct_present = n_present / n_assemb_neolithic,
    pct_gthalf = n_gthalf / n_assemb_neolithic,
    pct_gtqrtr = n_gtqrtr / n_assemb_neolithic
  ) %>% 
  arrange(-n_present, .by_group = TRUE) %>% 
  # slice_max(n_present, n = 10, with_ties = FALSE) %>%
  gt() %>% 
  fmt_percent(c(pct_present, pct_gthalf, pct_gtqrtr)) %>% 
  cols_merge_n_pct(n_present, pct_present) %>% 
  cols_merge_n_pct(n_gthalf, pct_gthalf) %>% 
  cols_merge_n_pct(n_gtqrtr, pct_gtqrtr) %>% 
  cols_label(taxon = "Taxon",
             n_present = "N (%) recorded present",
             n_gthalf = "N (%) making up more than half of assemblage",
             n_gtqrtr = "N (%) making up more than a quarter of assemblage") %>% 
  tab_style(cell_text(style = "italic"), cells_body(taxon)) ->
  occ_all

occ_all
# gtsave(occ_all, "figures/occ_all.html")
```

#### Founder crops by period

The following table shows the *occurrence* of the founder crop species in assemblages broken down by period, i.e. for each period, the number of assemblages where *N* distinct founder species are present, for varying values of *N*.

```{r occ-founder, layout='l-body-outset'}
flora %>% 
  mutate(founder_crop = recode(founder_crop,
                               "emmer wheat" = "wheat",
                               "einkorn wheat" = "wheat",
                               .default = founder_crop)) %>% 
  group_by(period, site_name, phase_code) %>% 
  summarise(n_founder_crop = n_distinct(founder_crop, na.rm = TRUE),
            .groups = "drop") %>% 
  group_by(period) %>% 
  summarise(
    n_assemb = n(),
    n_atleast7 = sum(n_founder_crop >= 7),
    p_atleast7 = n_atleast7 / n_assemb,
    n_atleast6 = sum(n_founder_crop >= 6),
    p_atleast6 = n_atleast6 / n_assemb,
    n_atleast5 = sum(n_founder_crop >= 5),
    p_atleast5 = n_atleast5 / n_assemb,
    n_atleast4 = sum(n_founder_crop >= 4),
    p_atleast4 = n_atleast4 / n_assemb,
    n_atleast3 = sum(n_founder_crop >= 3),
    p_atleast3 = n_atleast3 / n_assemb,
    n_atleast2 = sum(n_founder_crop >= 2),
    p_atleast2 = n_atleast2 / n_assemb,
    n_atleast1 = sum(n_founder_crop >= 1),
    p_atleast1 = n_atleast1 / n_assemb
  ) %>% 
  gt() %>% 
  fmt_percent(starts_with("p_")) %>% 
  cols_merge_n_pct(n_atleast7, p_atleast7) %>% 
  cols_merge_n_pct(n_atleast6, p_atleast6) %>% 
  cols_merge_n_pct(n_atleast5, p_atleast5) %>% 
  cols_merge_n_pct(n_atleast4, p_atleast4) %>% 
  cols_merge_n_pct(n_atleast3, p_atleast3) %>% 
  cols_merge_n_pct(n_atleast2, p_atleast2) %>% 
  cols_merge_n_pct(n_atleast1, p_atleast1) %>% 
  tab_spanner("Number of assemblages with N founder crops present:",
              columns = starts_with("n_atleast"), id = "n_atleast") %>% 
  cols_label(
    period = "Period",
    n_assemb = "Total assemblages",
    n_atleast7 = "All",
    n_atleast6 = ">= 6",
    n_atleast5 = ">= 5",
    n_atleast4 = ">= 4",
    n_atleast3 = ">= 3",
    n_atleast2 = ">= 2",
    n_atleast1 = ">= 1"
  ) %>% 
  # Doesn't work with cols_merge_n_pct apparently :(
  # data_color(
  #   columns = starts_with("p_atleast"),
  #   colors = scales::col_numeric(
  #     palette = paletteer::paletteer_d(
  #       palette = "ggsci::light_green_material"
  #       ) %>% as.character(),
  #     domain = c(0, 1)
  #     )
  # ) %>% 
  tab_footnote("Emmer wheat and einkorn wheat are counted as one crop.",
               locations = cells_column_spanners())
```

We can also look at the abundance of the founder crops – this time varying *P*, the proportion of the assemblage accounted for by the founder crops.

Here we exclude some records from the source databases (primarily ADEMNES) which do not actually contain quantitative information in the 'proportion' field, and instead use it to represent presence/absence.
We will do the same for following analyses that use abundance measures.

**TODO**

#### Founder crops by century

For a more fine-grained look at the time series, we can use the century bins calculated above, 
Measured by cross-assemblage ubiquity (number of assemblages dated to each century where the taxon is present) and by relative abundance (average proportion across assemblages dated to each century).

```{r ts-founder-crops, fig.height=6, layout='l-body-outset'}
# By ubiquity
flora_ts %>% 
  mutate(founder_crop = recode(founder_crop,
                               "einkorn wheat" = "wheat",
                               "emmer wheat" = "wheat",
                               # .missing = "other (not founder crop)",
                               .default = founder_crop)) %>% 
  count(century, founder_crop) %>% 
  mutate(founder_crop = fct_reorder(founder_crop, n, .fun = median)) %>% 
  drop_na(founder_crop) %>% 
  ggplot(aes(century / 1000, n, fill = founder_crop)) +
  geom_stream(sorting = "onset") +
  scale_x_reverse(breaks = scales::breaks_width(-2)) +
  khroma::scale_fill_muted() +
  labs(title = "Founder crops: occurrence & abundance by century",
       subtitle = "Cross-assemblage ubiquity",
       x = NULL, y = "n assemblages present", fill = NULL) +
  theme_minimal() ->
  p1

# By average proportion
flora_ts %>%
  group_by(site_name, phase_code) %>%
  filter(sum(prop, na.rm = TRUE) != 0) %>% # Exclude site-phases with no quantitative data
  ungroup() %>%
  mutate(founder_crop = recode(founder_crop,
                               "einkorn wheat" = "wheat",
                               "emmer wheat" = "wheat",
                               # .missing = "other (not founder crop)",
                               .default = founder_crop)) %>% 
  group_by(century, site_name, phase_code, founder_crop) %>% 
  summarise(prop = sum(prop, na.rm = TRUE), .groups = "drop") %>% 
  group_by(century) %>% 
  mutate(n_assemb = length(unique(phase_code))) %>% 
  ungroup() %>% 
  group_by(century, founder_crop) %>%
  summarise(avg_prop = sum(prop) / first(n_assemb)) %>%
  ggplot(aes(century / 1000, avg_prop, fill = founder_crop)) +
  geom_stream(type = "ridge", sorting = "onset") +
  scale_x_reverse(breaks = scales::breaks_width(-2)) +
  scale_y_continuous(limits = c(0, 1), breaks = scales::breaks_pretty(5),
                     labels = scales::percent) +
  khroma::scale_fill_muted() +
  labs(subtitle = "Relative abundance",
       x = "ka cal BP", y = "average proportion", fill = NULL) +
  theme_minimal() ->
  p2

p1 / p2

ggsave(here("analysis/figures/ts-founder-crops.pdf"), width = 160, height = 160,
       units = "mm")
```

#### Grasses
 
#### Legumes

#### Wild plants

#### Fruits and nuts

### Were the founder crops the *earliest* plants?

Our discussion of the earliest occurrences of certain taxa was informed by the following interrogations of our combined flora dataset.

#### *T. aestivum/durum.*

Earliest reported occurrences:

```{r occ-t-aestivum-durum, layout='l-body-outset'}
flora %>% 
  filter(taxon_detail %in% c("Triticum aestivum", "Triticum durum") | 
           taxon_source %in% c("Triticum aestivum/durum",
                               "Triticum dicoccum/aestivum",
                               "Triticum dicoccum/durum")) %>% 
  arrange(-age_start) %>% 
  group_by(source, site_name, taxon_source) %>% 
  summarise(phase_code = first(phase_code), earliest_age = first(age_start),
            n = sum(n), reference = first(reference),
            .groups = "drop") %>% 
  mutate(reference = str_remove(reference, ";.*")) %>% 
  arrange(-earliest_age) %>% 
  slice(1:10) %>% 
  gt() %>% 
  cols_label(source = "Database", site_name = "Site", phase_code = "Phase",
             taxon_source = "Reported taxon", 
             earliest_age = "Reported age (earliest)", n = "N",
             reference = "Reference")
```

#### *T. spelta*

Rarely reported, so all sites it occurs at, ordered by age:

```{r occ-t-spelta, layout='l-body-outset'}
flora %>% 
  filter(taxon_detail == "Triticum spelta" | 
           taxon_source %in% c("Triticum dicoccum/spelta",
                               "Triticum dicoccum/spelta grains")) %>% 
  arrange(-age_start) %>% 
  group_by(source, site_name, taxon_source) %>% 
  summarise(phase_code = first(phase_code), earliest_age = first(age_start),
            n = sum(n), reference = first(reference),
            .groups = "drop") %>% 
  mutate(reference = str_remove(reference, ";.*")) %>% 
  arrange(-earliest_age) %>% 
  slice(1:10) %>% 
  gt() %>% 
  cols_label(source = "Database", site_name = "Site", phase_code = "Phase",
             taxon_source = "Reported taxon", 
             earliest_age = "Reported age (earliest)", n = "N",
             reference = "Reference")
```

#### *T. monococcum*

Earliest reported assemblages with >5%, from the LPPNB/C or later:

```{r occ-monococcum, layout='l-body-outset'}
flora %>% 
  filter(taxon_detail == "Triticum monococcum") %>% 
  filter(prop > 0.05) %>% 
  filter(age_start < periods$start_bp[periods$period == "LPPNB/C"]) %>% 
  arrange(-age_start) %>% 
  group_by(source, site_name, taxon_source) %>% 
  summarise(phase_code = first(phase_code), earliest_age = first(age_start),
            prop = max(prop), reference = first(reference),
            .groups = "drop") %>% 
  mutate(reference = str_remove(reference, ";.*")) %>% 
  arrange(-earliest_age) %>% 
  slice(1:10) %>% 
  gt() %>% 
  cols_label(source = "Database", site_name = "Site", phase_code = "Phase",
             taxon_source = "Reported taxon", 
             earliest_age = "Reported age (earliest)", prop = "Highest %",
             reference = "Reference") %>% 
  fmt_percent(vars(prop))
```

#### *H. spontaneum/vulgare*

Earliest reported assemblages with >5%:

```{r occ-hordeum, layout='l-body-outset'}
flora %>% 
  filter(taxon_detail %in% c("Hordeum spontaneum", "Hordeum vulgare")) %>% 
  filter(prop > 0.05) %>% 
  arrange(-age_start) %>% 
  group_by(source, site_name, taxon_source) %>% 
  summarise(phase_code = first(phase_code), earliest_age = first(age_start),
            prop = max(prop), reference = first(reference),
            .groups = "drop") %>% 
  mutate(reference = str_remove(reference, ";.*")) %>% 
  arrange(-earliest_age) %>% 
  slice(1:10) %>% 
  gt() %>% 
  cols_label(source = "Database", site_name = "Site", phase_code = "Phase",
             taxon_source = "Reported taxon", 
             earliest_age = "Reported age (earliest)", prop = "Highest %",
             reference = "Reference") %>% 
  fmt_percent(vars(prop))
```

#### *Avena*

Earliest reported assemblages with >2% (there aren't many):

```{r occ-avena, layout='l-body-outset'}
flora %>% 
  filter(genus == "Avena") %>% 
  filter(prop > 0.02) %>% 
  arrange(-age_start) %>% 
  group_by(source, site_name, taxon_source) %>% 
  summarise(phase_code = first(phase_code), earliest_age = first(age_start),
            prop = max(prop), reference = first(reference),
            .groups = "drop") %>% 
  mutate(reference = str_remove(reference, ";.*")) %>% 
  arrange(-earliest_age) %>% 
  slice(1:10) %>% 
  gt() %>% 
  cols_label(source = "Database", site_name = "Site", phase_code = "Phase",
             taxon_source = "Reported taxon", 
             earliest_age = "Reported age (earliest)", prop = "Highest %",
             reference = "Reference") %>% 
  fmt_percent(vars(prop))
```

#### *Linum usitatissimum*

Earliest reported occurrences of *Linum usitatissimum* (flax) specifically:

```{r occ-flax, layout='l-body-outset'}
flora %>% 
  filter(taxon_detail == "Linum usitatissimum" | 
           taxon_source == "Linum bienne/usitatissimum") %>% 
  arrange(-age_start) %>% 
  group_by(source, site_name, taxon_source) %>% 
  summarise(phase_code = first(phase_code), earliest_age = first(age_start),
            n = sum(n), reference = first(reference),
            .groups = "drop") %>% 
  mutate(reference = str_remove(reference, ";.*")) %>% 
  arrange(-earliest_age) %>% 
  slice(1:10) %>% 
  gt() %>% 
  cols_label(source = "Database", site_name = "Site", phase_code = "Phase",
             taxon_source = "Reported taxon", 
             earliest_age = "Reported age (earliest)", n = "N",
             reference = "Reference")
```

